**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-10-20 paper/manuscript.Rmd
remove hyphens, add running title, update figS4

---
bibliography: references.bib
csl: msystems.csl
output:
  rmarkdown::pdf_document:
    keep_tex: yes
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{multirow}
---

@@

@@

### Abstract
Colorectal cancer is a common and deadly disease in the United States accounting for over 50,000 deaths in 2020. This progressive disease is highly preventable with early detection and treatment, but many people do not comply with the recommended screening guidelines. The gut microbiome has emerged as a promising target for non-invasive detection of colorectal cancer. Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs) with the goal of increasing taxonomic resolution. However, it is unknown which taxonomic resolution is optimal for microbiome-based classification of colorectal cancer. To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels. We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p > 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p < 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`). These results demonstrate a trade-off between taxonomic resolution and prediction performance, where coarse taxonomic resolution (e.g. phylum) is not distinct enough but fine resolution (e.g. ASV) is to individualized to accurately classify samples. Similar to the story of Goldilocks and the three bears, mid-range resolution (i.e. family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.  
(238 words)

### Importance
Despite being highly preventable, colorectal cancer remains a leading cause of cancer related death in the United States. Low-cost, non-invasive detection methods could greatly improve our ability to identify and treat early stages of disease. The microbiome has shown promise as a resource for detection of colorectal cancer. Research on the gut microbiome tends to focus on improving our ability to profile species and strain level taxonomic resolution. However, we found that finer resolution impedes the ability to predict colorectal cancer based on the gut microbiome. These results highlight the need for consideration of the appropriate taxonomic resolution for microbiome analyses and that finer resolution is not always more informative.  
(110 words)

>>>>

---
bibliography: references.bib
csl: msystems.csl
output:
  rmarkdown::pdf_document:
    keep_tex: yes
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{multirow}
  - \usepackage[none]{hyphenat}
---

@@

**Running title:** A Goldilocks Principle for the Gut Microbiome

@@

### Abstract
Colorectal cancer is a common and deadly disease in the United States accounting for over 50,000 deaths in 2020. This progressive disease is highly preventable with early detection and treatment, but many people do not comply with the recommended screening guidelines. The gut microbiome has emerged as a promising target for non-invasive detection of colorectal cancer. Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs) with the goal of increasing taxonomic resolution. However, it is unknown which taxonomic resolution is optimal for microbiome-based classification of colorectal cancer. To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels. We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p > 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p < 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`). These results demonstrate a trade-off between taxonomic resolution and prediction performance, where coarse taxonomic resolution (e.g. phylum) is not distinct enough but fine resolution (e.g. ASV) is to individualized to accurately classify samples. Similar to the story of Goldilocks and the three bears, mid-range resolution (i.e. family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.

### Importance
Despite being highly preventable, colorectal cancer remains a leading cause of cancer related death in the United States. Low-cost, non-invasive detection methods could greatly improve our ability to identify and treat early stages of disease. The microbiome has shown promise as a resource for detection of colorectal cancer. Research on the gut microbiome tends to focus on improving our ability to profile species and strain level taxonomic resolution. However, we found that finer resolution impedes the ability to predict colorectal cancer based on the gut microbiome. These results highlight the need for consideration of the appropriate taxonomic resolution for microbiome analyses and that finer resolution is not always more informative.

### 2021-10-20 paper/manuscript.tex
remove hyphens, add running title, update figS4

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

@@

Colorectal cancer is a common and deadly disease in the United States
accounting for over 50,000 deaths in 2020. This progressive disease is
highly preventable with early detection and treatment, but many people
do not comply with the recommended screening guidelines. The gut
microbiome has emerged as a promising target for non-invasive detection
of colorectal cancer. Most microbiome-based classification efforts
utilize taxonomic abundance data from operational taxonomic units (OTUs)
or amplicon sequence variants (ASVs) with the goal of increasing
taxonomic resolution. However, it is unknown which taxonomic resolution
is optimal for microbiome-based classification of colorectal cancer. To
address this question, we used a reproducible machine learning framework
to quantify classification performance of models based on data annotated
to phylum, class, order, family, genus, OTU, and ASV levels. We found
that model performance increased with increasing taxonomic resolution,
up to the family level where performance was equal (p \textgreater{}
0.05) among family (mean AUC: 0.689), genus (mean AUC: 0.690), and OTU
(mean AUC: 0.693) levels before decreasing at the ASV level (p
\textless{} 0.05, mean AUC: 0.676). These results demonstrate a
trade-off between taxonomic resolution and prediction performance, where
coarse taxonomic resolution (e.g.~phylum) is not distinct enough but
fine resolution (e.g.~ASV) is to individualized to accurately classify
samples. Similar to the story of Goldilocks and the three bears,
mid-range resolution (i.e.~family, genus, and OTU) is just right for
optimal prediction of colorectal cancer from microbiome data.\\
(238 words)

@@

Despite being highly preventable, colorectal cancer remains a leading
cause of cancer related death in the United States. Low-cost,
non-invasive detection methods could greatly improve our ability to
identify and treat early stages of disease. The microbiome has shown
promise as a resource for detection of colorectal cancer. Research on
the gut microbiome tends to focus on improving our ability to profile
species and strain level taxonomic resolution. However, we found that
finer resolution impedes the ability to predict colorectal cancer based
on the gut microbiome. These results highlight the need for
consideration of the appropriate taxonomic resolution for microbiome
analyses and that finer resolution is not always more informative.\\
(110 words)

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\usepackage{multirow}
\usepackage[none]{hyphenat}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\textbf{Running title:} A Goldilocks Principle for the Gut Microbiome

@@

Colorectal cancer is a common and deadly disease in the United States
accounting for over 50,000 deaths in 2020. This progressive disease is
highly preventable with early detection and treatment, but many people
do not comply with the recommended screening guidelines. The gut
microbiome has emerged as a promising target for non-invasive detection
of colorectal cancer. Most microbiome-based classification efforts
utilize taxonomic abundance data from operational taxonomic units (OTUs)
or amplicon sequence variants (ASVs) with the goal of increasing
taxonomic resolution. However, it is unknown which taxonomic resolution
is optimal for microbiome-based classification of colorectal cancer. To
address this question, we used a reproducible machine learning framework
to quantify classification performance of models based on data annotated
to phylum, class, order, family, genus, OTU, and ASV levels. We found
that model performance increased with increasing taxonomic resolution,
up to the family level where performance was equal (p \textgreater{}
0.05) among family (mean AUC: 0.689), genus (mean AUC: 0.690), and OTU
(mean AUC: 0.693) levels before decreasing at the ASV level (p
\textless{} 0.05, mean AUC: 0.676). These results demonstrate a
trade-off between taxonomic resolution and prediction performance, where
coarse taxonomic resolution (e.g.~phylum) is not distinct enough but
fine resolution (e.g.~ASV) is to individualized to accurately classify
samples. Similar to the story of Goldilocks and the three bears,
mid-range resolution (i.e.~family, genus, and OTU) is just right for
optimal prediction of colorectal cancer from microbiome data.

@@

Despite being highly preventable, colorectal cancer remains a leading
cause of cancer related death in the United States. Low-cost,
non-invasive detection methods could greatly improve our ability to
identify and treat early stages of disease. The microbiome has shown
promise as a resource for detection of colorectal cancer. Research on
the gut microbiome tends to focus on improving our ability to profile
species and strain level taxonomic resolution. However, we found that
finer resolution impedes the ability to predict colorectal cancer based
on the gut microbiome. These results highlight the need for
consideration of the appropriate taxonomic resolution for microbiome
analyses and that finer resolution is not always more informative.
